### 2023-02-27 calculus.tex
Draft mechanization section

\section{Marked Lambda Calculus}
\label{sec:calculus}

@@

Again, to satisfy totality, we must consider the case when $\TMV$ is not a matched arrow type. There
is no expected type for the argument, so we perform analytic marking on $\EMV_2$ against the
unknown type. The applied expression is placed inside the inconsistent type mark, which is in
accordance with the meaning of that mark.
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV} \\
    \notMatchedArrow{\TMV} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TUnknown}
  }{
    \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECApNonMatched{\ECMV_1}{\ECMV_2}}{\TUnknown}
  }{ISAp2}

@@

\begin{figure}[htbp]
  \raggedright
  \judgbox{\ensuremath{\TJoin{\TMV_1}{\TMV_2}}} is a \emph{partial} metafunction defined as follows:
  %
  \newcommand{\joinsTo}[3]{\ensuremath{\TJoin{#1}{#2} & = & #3}}
  \[\begin{array}{rcl}
    \joinsTo{\TUnknown}{\TMV}{\TMV} \\
    \joinsTo{\TMV}{\TUnknown}{\TMV} \\
    \joinsTo{\TNum}{\TNum}{\TNum} \\
    \joinsTo{\TBool}{\TBool}{\TBool} \\
    \joinsTo{(\TArrow{\TMV_1}{\TMV_2})}{(\TArrow{\TMV_1'}{\TMV_2'})}{\TArrow{(\TJoin{\TMV_1}{\TMV_1'})}{(\TJoin{\TMV_2}{\TMV_2'})}} \\
  \end{array}\]
  %
  \caption{Type join.}
  \label{fig:calculus-type-join}
\end{figure}
% TODO intuition for type join?

@@

Finally, we consider the subsumption rule in typing unmarked expressions, which states that if an
expression synthesizes a type, it may also be analysed against that type or any consistent type.
\[%
  \judgment{
    \ctxSynType{\ctx}{\EMV}{\TMV'} \\
    \consistent{\TMV}{\TMV'} \\
    \subsumable{\EMV}
  }{
    \ctxAnaType{\ctx}{\EMV}{\TMV}
  }{UASubsume}
\]%
We also restrict the usage of subsumption to ``subsumable'' syntactic forms, written
$\subsumable{\EMV}$. This judgment is defined on all syntactic forms except lambda abstractions and
conditionals, the only ones with explicit analysis rules. In other words, we restrict
subsumption to be the analytic rule of ``last resort''. This is critical in establishing that
marking and typing are deterministic (discussed later in\cref{thm:calculus-marking-unicity}).

@@

\begin{theorem}[name=Marking of Well-Typed/Ill-Typed Expressions] \ % TODO better name for this
  \label{thm:calculus-marking-well-ill-typed}
  \begin{enumerate}
    \item \begin{enumerate}
        \item If $\ctxSynType{\ctx}{\EMV}{\TMV}$, then $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$
          such that $\imageIs{\EMV}{\ECMV}$.
        \item If $\ctxAnaType{\ctx}{\EMV}{\TMV}$, then $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$
          such that $\imageIs{\EMV}{\ECMV}$.
      \end{enumerate}

@@

We now demonstrate how one might extend or adapt the marking calculus for new typing features by
adding binary product types. First, we extend in \cref{fig:calculus-products-judgments} the syntax
of types and notions of consistency and join. As with arrow types, we additionally define an
auxiliary notion of \emph{matched binary product types}: the judgment
$\matchedPair{\TMV}{\TMV_1}{\TMV_2}$ establishes that $\TMV$ may be considered the binary product
type $\TPair{\TMV_1}{\TMV_2}$.

\input{figs/calculus-products-judgments}

The unmarked language with a pair constructor and projection operators.
\Cref{fig:calculus-products-uexp} gives their syntax and typing rules. Similarly to application,
\textsc{USProjL} and \textsc{USProjR} employ the matched binary product type; a projection
$\EProjL{\EMV}$ or $\EProjR{\EMV}$ is then well-typed if the $\EMV$ may be considered to be a pair.

\input{figs/calculus-products-uexp}

@@

@@

The metatheory of the calculus has been fully mechanized in the Agda proof assistant [citation]. The
mechanization's documentation contains more details regarding technical decisions made therein.

>>>>

\section{The Marking Calculus}
\label{sec:calculus}

@@

Again, to satisfy totality, we must consider the case when $\TMV$ is not a matched arrow type, such
as in the example of \cref{fig:calculus-examples-app-non-lambda}. There is no expected type for the
argument, so we perform analytic marking on $\EMV_2$ against the unknown type. The applied
expression is placed inside the inconsistent type mark, which is in accordance with the meaning of
that mark.
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV} \\
    \notMatchedArrow{\TMV} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TUnknown}
  }{
    \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECApNonMatched{\ECMV_1}{\ECMV_2}}{\TUnknown}
  }{ISAp2}

@@

\newcommand{\joinsTo}[3]{\ensuremath{\TJoin{#1}{#2} & = & #3}}
\begin{figure}[htbp]
  \raggedright
  \judgbox{\ensuremath{\TJoin{\TMV_1}{\TMV_2}}} is a \emph{partial} metafunction defined as follows:
  %
  \[\begin{array}{rcl}
    \joinsTo{\TUnknown}{\TMV}{\TMV} \\
    \joinsTo{\TMV}{\TUnknown}{\TMV} \\
    \joinsTo{\TNum}{\TNum}{\TNum} \\
    \joinsTo{\TBool}{\TBool}{\TBool} \\
    \joinsTo{(\TArrow{\TMV_1}{\TMV_2})}{(\TArrow{\TMV_1'}{\TMV_2'})}{\TArrow{(\TJoin{\TMV_1}{\TMV_1'})}{(\TJoin{\TMV_2}{\TMV_2'})}} \\
  \end{array}\]
  %
  \caption{Type join.}
  \label{fig:calculus-type-join}
\end{figure}
% TODO intuition for type join?

@@

Finally, we consider the subsumption rule in typing unmarked expressions, which states that if an
expression synthesizes a type, it may also be analysed against that type or any consistent type.
\[%
  \judgment{
    \ctxSynType{\ctx}{\EMV}{\TMV'} \\
    \consistent{\TMV}{\TMV'} \\
    \subsumable{\EMV}
  }{
    \ctxAnaType{\ctx}{\EMV}{\TMV}
  }{UASubsume}
\]%
We also restrict the usage of subsumption to ``subsumable'' syntactic forms, written
$\subsumable{\EMV}$. This judgment is defined on all syntactic forms except lambda abstractions and
conditionals, the only ones with explicit analysis rules. In other words, we restrict
subsumption to be the analytic rule of ``last resort''. This is critical in establishing that
marking and typing are deterministic (discussed later in \cref{thm:calculus-marking-unicity}).

@@

\begin{theorem}[name=Marking of Well-Typed/Ill-Typed Expressions] \ % TODO better name for this
  \label{thm:calculus-marking-well-ill-typed}
  \begin{enumerate}
    \item \begin{enumerate}
        \item If $\ctxSynType{\ctx}{\EMV}{\TMV}$ and $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$,
          then $\imageIs{\EMV}{\ECMV}$.
        \item If $\ctxAnaType{\ctx}{\EMV}{\TMV}$ and $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$,
          then $\imageIs{\EMV}{\ECMV}$.
      \end{enumerate}

@@

We now demonstrate how one might extend or adapt the marking calculus for new typing features by
adding binary product types. First, we extend the syntax of types and the accompanying notions of
consistency and join, which closely mirror the definitions for arrow types.
%
\[\begin{array}{rrcl}
  \TMName  & \TMV  & \Coloneqq & \cdots \mid \TPair{\TMV}{\TMV}
\end{array}\]

We additionally define an auxiliary notion of \emph{matched binary product types} in
\cref{fig:calculus-matched-product}: the judgment $\matchedPair{\TMV}{\TMV_1}{\TMV_2}$ establishes
that $\TMV$ may be considered the binary product type $\TPair{\TMV_1}{\TMV_2}$.

\begin{figure}[htbp]
  \raggedright
  \judgbox{\ensuremath{\matchedPair{\TMV}{\TMV_1}{\TMV_2}}} $\TMV$ has matched binary product type $\TPair{\TMV_1}{\TMV_2}$
  %
  \begin{mathpar}
    \judgment{ }{
      \matchedPair{\TUnknown}{\TUnknown}{\TUnknown}
    }{TMPHole}

    \judgment{ }{
      \matchedPair{\TPair{\TMV_1}{\TMV_2}}{\TMV_1}{\TMV_2}
    }{TMPProd}
  \end{mathpar}
  %
  \caption{Matched product type.}
  \label{fig:calculus-matched-product}
\end{figure}

The unmarked language is extended with a pair constructor and projection operators.
%
\[\begin{array}{rrcl}
  \EMName  & \EMV  & \Coloneqq & \cdots
                               \mid \EPair{\EMV}{\EMV}
                               \mid \EProjL{\EMV} \mid \EProjR{\EMV}
\end{array}\]
%
\begin{mathpar}
  \judgment{
    \ctxSynType{\ctx}{\EMV_1}{\TMV_1} \\
    \ctxSynType{\ctx}{\EMV_2}{\TMV_2}
  }{
    \ctxSynType{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\TPair{\TMV_1}{\TMV_2}}
  }{USPair}

  \judgment{
    \ctxSynType{\ctx}{\EMV}{\TMV} \\
    \matchedPair{\TMV}{\TMV_1}{\TMV_2}
  }{
    \ctxSynType{\ctx}{\EProjL{\EMV}}{\TMV_1}
  }{USProjL}

  \judgment{
    \ctxSynType{\ctx}{\EMV}{\TMV} \\
    \matchedPair{\TMV}{\TMV_1}{\TMV_2}
  }{
    \ctxSynType{\ctx}{\EProjR{\EMV}}{\TMV_2}
  }{USProjR} \\
\end{mathpar}
%
Similarly to application, \textsc{USProjL} and \textsc{USProjR} employ the matched binary product
type; a projection $\EProjL{\EMV}$ or $\EProjR{\EMV}$ is then well-typed if the $\EMV$ may be
considered to be a pair.

@@

We now consider how the calculus might be extended to richer typing features such as parametric
polymorphism. Specifically, we consider System F-style explicit polymorphism.

@@

The semantics and metatheory presented above, including the extension with binary products, have
been fully mechanized in the Agda proof assistant [citation]. Though the mechanization's
documentation contains more detailed discussion regarding technical decisions made therein, we
highlight some important aspects here.

The standard approach of modeling judgments as inductive datatypes and rules as constructors for
those datatypes is taken. For marked expressions, we opt for implicitly typed terms [citation],
which gives us part of \cref{thm:calculus-marking-well-formedness} for free. For the convenience of
readers interested in browsing the mechanization, all rule names match those presented here.

\section{Marking Calculus in Structured Editing}
\label{sec:calculus-structured-editing}

The previous section presented a formal calculus for error localization and recovery. We now discuss
its usage in a structured editing setting.

In particular, the calculus has been integrated into the Hazel programming environment, a live
programming environment with support for typed holes.
